---
title: "Governors: Wisdom and Justice"
format:
  html: default
execute:
  echo: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(tidymodels)
library(marginaleffects)

governors <- read_csv("data/govenors.csv", show_col_types = FALSE)
```

## The Question

There are two scenarios. Focus on just one of them:

- Imagine you work for a life insurer, and want to forecast how long candidates for the US Senate might live based on their age, party, election result, and other variables.

- Imagine you are a researcher. You want to know if winning candidates live longer.

The life insurer asks: How many years, on average, will this year's US Senate candidates live after their election, given their age, party, and election result?

The researcher asks: What is the average causal effect of winning an election, rather than losing it, on the number of years a candidate lives afterward?

## Wisdom

### Preceptor Table

**What are the units, outcome, and covariates for your scenario's question?**

- Units:

The U.S Senate Cantidates

- Outcomes:

How long the Cantidates live after their election day.

- Covariates:

(Could include) Age, party, gender, and election results.

- Treatment (if any):

None (cause this is a predictive question)

- Quantity of interest:

The expected years of each cantidate lives (given covariates)

Describe the key components of a Preceptor Table for your scenario; Use words like "units," "outcomes," and "covariates."



### Exploratory Data Analysis

Create a plot that shows the distribution of `lived_after`.

```{r}
#| label: lived-after-distribution
#| message: false

governors |> 
  ggplot(aes(x = lived_after)) +
  geom_histogram(binwidth = 5, color = "white", fill = "dodgerblue4") +
  labs(
    title = "Distribution of Years Lived After Election",
    x = "Years Lived After Election",
    y = "Count"
  ) +
  theme_minimal()
```

Create a tibble, called `x`, which keeps only close elections (`win_margin` within 5 points) after 1945 and adds a variable `election_result` recording whether the candidate won or lost. How many candidates remain? Compare the average `lived_after`, and the average `election_age`, of winners and losers.

```{r}
#| label: close-elections-subset
#| message: false

x <- governors |> 
  filter(abs(win_margin) <= 5, year > 1945) |> 
  mutate(election_result = if_else(outcome == "Win", "won", "lost"))

x |> 
  group_by(election_result) |> 
  summarize(
    count = n(),
    avg_lived_after = mean(lived_after, na.rm = TRUE),
    avg_election_age = mean(election_age, na.rm = TRUE)
  )
```

There are **254** candidates remaining in the subset. On average, winners lived **29.6** years after their election (with an average election age of **51.0** years), whereas losers lived **27.2** years after their election (with an average election age of **51.4** years).


Using `x`, create a plot that shows the relationship between `lived_after` and `election_age`.

```{r}
#| label: lived-after-vs-age
#| message: false

x |> 
  ggplot(aes(x = election_age, y = lived_after, color = election_result)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(
    title = "Life Expectancy Post-Election vs. Age at Election",
    subtitle = "Close Gubernatorial Elections after 1945",
    x = "Age at Election (years)",
    y = "Years Lived After Election",
    color = "Election Result"
  ) +
  scale_color_manual(values = c("won" = "dodgerblue4", "lost" = "firebrick")) +
  theme_minimal()
```

------------------------------------------------------------------------

## Justice

### Population Table

Describe the key components of a Population Table for your scenario.

The Population Table would have the data from the Preceptors Tables along with the data we actually have.

### Assumptions

Why did we restrict our data to close elections?

So things remain unconfounded. The treatment of winning or losing is applied essentially at random.

Counter-example to the assumption of validity:

If the preceptors table is regarding senate elections but our data is for governer elections. Winning a senator election has a different meaning than winning a governer election. 

Counter-example to the assumption of stability:

Holding office in earlier years could have a very different affect on life expectancy than the present.

Counter-example to the assumption of representativeness:

If there's a gender imbalance in the past that isn't very representative of the future

Counter-example to the assumption of unconfoundedness:

None because our question is predictive.

### Probability Family and Link Function

What probability family should we use?



Why should we use that probability family?



What is the link function for our outcome?
